An Aliasing-Tolerant Parallel Programming Model

### 2012-11-11 doc/types-pldi2013/langdef.tex
Lots of small changes and moving figures around
Fixing some typesetting issues for inline math modes

To determine when parallel execution is safe, we start from the sequential execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
traces will be equivalent to a sequential execution and parallel execution is safe.  To
efficiently discover these cases at runtime, we require a test that can determine this property prior
to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and 
the current region mapping can soundly predict when this property will hold.
%If we can show that reordering is allowed
%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
%privileges and the current region mapping can soundly predict when this property will hold.
We begin by defining a {\em non-interference} operator on two memory operations
$\epsilon_1 = op_1(l_1, c_1, v_1)$ and $\epsilon_2 = op_2(l_2, c_2, v_2)$:
$$
\begin{array}{l@{ }l}
\epsilon_1\ \nonint\ \epsilon_2 \Leftrightarrow & (op_1 = read \wedge op_2 = read)\ \vee \\
& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \vee \\
& l_1 \not= l_2
\end{array}
$$
Reads have no side effects, and cannot change what another read returns.  The safety of the second case
follows from the requirement that reduction operations be commutative.  
Finally, accesses to different locations cannot affect each other.
We next show that this non-interference operator preserves the validity of an interleaving
of two or more traces when reordering a pair of memory operations.
\begin{lem}
\label{lem:nonintswap}
\rm
Let $S$ be a store, $C$ a clobber set, $\oton{E}, E'_a, E'_b$ memory traces, and
$\epsilon_1, \epsilon_2$ be two memory operations from $E_i$ and $E_j$ ($i \not= j$).  Then,
$$\begin{array}{l@{}l}
valid\_interleave&(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E}) \wedge \epsilon_1 \nonint \epsilon_2 \Rightarrow \\
 & valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).
\end{array}$$
\end{lem}
%The proof follows from the definitions of {\em coherent} and {\em seq\_equiv} in Figure~\ref{fig:validinterleave}.

@@

In cases where Legion cannot safely infer non-interference of
privileges (perhaps because two tasks actually access the same data in aliased regions),
relaxation of the constraints on execution order can
still be requested by the programmer through the use of coherence 
annotations on individual regions passed to a task.
%the default behavior of sequential execution can
%be avoided through the use of relaxed coherence.
The {\em atomic} coherence mode specifies that although two
tasks interfere due to accessing aliased regions, they may execute in either order, allowing
the task issued later in program order to possibly run before the task issued earlier 
in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
%or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).
To show this is safe, we define a relaxed version of non-interference for atomic coherence:
$$
\begin{array}{l@{}l}
op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint\ op_2(l_2, c_2, v_2) \vee \\
& (c_1 = atomic \wedge c_2 = atomic)
\end{array}$$

>>>>

To determine when parallel execution is safe, we start from the sequential execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
traces will be equivalent to a sequential execution and parallel execution is safe.  To
efficiently discover these cases at runtime, we require a test that can determine this property prior
to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and 
the current region mapping can soundly predict when this property will hold.
%If we can show that reordering is allowed
%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
%privileges and the current region mapping can soundly predict when this property will hold.
We begin by defining a {\em non-interference} operator on two memory operations
$\epsilon_1 = op_1(l_1, c_1, v_1)$ and $\epsilon_2 = op_2(l_2, c_2, v_2)$:
$$
\begin{array}{l@{ }l}
\epsilon_1\ \nonint\ \epsilon_2 \Leftrightarrow & (op_1 = read \wedge op_2 = read)\ \vee \\
& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \vee \\
& l_1 \not= l_2
\end{array}
$$
Reads have no side effects, and cannot change what another read returns.  The safety of the second case
follows from the requirement that reduction operations be commutative.  
Finally, accesses to different locations cannot affect each other.
We next show that this non-interference operator preserves the validity of an interleaving
of two or more traces when reordering a pair of memory operations.
\begin{lem}
\label{lem:nonintswap}
\rm
Let $S$ be a store, $C$ a clobber set, $\oton{E}, E'_a, E'_b$ memory traces, and
$\epsilon_1, \epsilon_2$ be two memory operations from $E_i$ and $E_j$ ($i \not= j$).  Then,
%%$$\begin{array}{l}
\begin{align*}
valid&\_interleave(S, C, E'_a \dplus [ \epsilon_1, \epsilon_2 ] \dplus E'_b, \oton{E}) \wedge \epsilon_1 \nonint \epsilon_2 \\
&\Rightarrow valid\_interleave(S, C, E'_a \dplus [ \epsilon_2, \epsilon_1 ] \dplus E'_b, \oton{E}).
\end{align*}
%%\end{array}$$
\end{lem}
%The proof follows from the definitions of {\em coherent} and {\em seq\_equiv} in Figure~\ref{fig:validinterleave}.

@@

In cases where Legion cannot safely infer non-interference of
privileges (perhaps because two tasks actually access the same data in aliased regions),
relaxation of the constraints on execution order can
still be requested by the programmer through the use of coherence 
annotations on individual regions passed to a task.
%the default behavior of sequential execution can
%be avoided through the use of relaxed coherence.
The {\em atomic} coherence mode specifies that although two
tasks interfere due to accessing aliased regions, they may execute in either order, allowing
the task issued later in program order to possibly run before the task issued earlier 
in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
%or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).
To show this is safe, we define a relaxed version of non-interference for atomic coherence:
%$$
%\begin{array}{l@{}l}
\begin{align*}
op_1 &(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow  \\
& op_1(l_1, c_1, v_1)\ \nonint\ op_2(l_2, c_2, v_2) \vee (c_1 = atomic \wedge c_2 = atomic)
\end{align*}
%\end{array}
%$$
